One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-08-02 papers/1_dmatpol/paper-r3.tex
Ikhusia's comments added

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities (DMS)
interacting with the spatial distribution of electric field.
The analytical approximate expressions were derived
at the Hartree\hyp{}Fock theory, which served as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron density of water molecule in spatially uniform electric field, as well as
in spatially non\hyp{}uniform electric field distribution
generetated by point charges. When both linear and quadratic terms with respect to the electric field 
are accounted for,
the electric field\hyp{}induced polarization energies,
dipole moments and quadrupole moments are quantitatively described by the present theory 
in electric fields from weak to very strong in magnitude (0.003 -- 0.07 a.u.).
It is believed that the proposed model could open
new routes in Quantum Chemistry of fast and efficient calculations of strongly perturbed systems.
\end{abstract}

@@

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,\cite{Holas.March.PhysRevA.1991}
though obtaining the exact density has been an undersolved issue for most of chemical systems.
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.

The simplest, yet perhaps the most relevant for chemistry, type of perturbation 
is the electrostatic perturbation.
It is worth mentioning that, at that moment, there is no simple \emph{ab initio}
method of computation of electron deformation densities without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Head-Gordon.JCP.2015,
Mandado.Hermida-Ramon.JCTC.2011}
In this work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element will accumulate when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let the induced dipole moment be partitioned into separate contributions
from the occupied MO's, i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarizability tensor
associated with the $i$th MO whereas ${\bf F}({\bf r}_i)$ is the electric field evaluated at the $i$th MO centroid of charge,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
We can now insert the above result into Eq.~\eqref{e:dmu-l-vector} 
and after recasting the right hand side in a matrix form
and assuming real quantities
we get
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of size $(n \times 3)$ and $\delta{\bf c}_i$ is just the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   \;,
\end{equation}
%
and has size $(3\times n)$. Note that the \emph{right inverse} of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic (and higher\hyp{}order) terms with respect to the electric field
can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \\
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\} \;.
\end{multline}
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is a $(3\times 3)$ matrix. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.\footnote{More general
contribution due to square of electric field could 
be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. 
However, for simplicity, we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
}

@@

The exact mathematical form of the generalized density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown. 
The susceptibility from Eq.~\eqref{e:final-model.HF} is valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarizability approximation), 
but not higher multipole moments and polarization energy, are correctly described.
Our strategy is to treat the susceptibilities from Eq.~\eqref{e:final-model.General}
as adjustable parameters, rather than try to derive them from first principles theory.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule
was selected as a test system which was subject to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields. 
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval of $\vert 0.0404 \vert$ a.u.
generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of perturbing charges were drawn from the uniform distribution 
of a set of fourty three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly).
In this way, independent sets of 100 random samples of uniform electric fields as well as 
40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.

@@

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;. 
  \end{align}
\end{subequations}
%
where the subscript `CM' denotes the center of mass of water molecule.
When the distributed polarizabilities were used instead of the polarization\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments (given in Eq.~\eqref{e:dmu-dpol}) according to
%
\begin{equation}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
Methodology for the calculation of the electric field due to point charges
can be found elsewhere.\cite{Blasiak.Cho.JCP.2014}

>>>>

The electric field\hyp{}induced change in the one\hyp{}electron density
has been expressed as a series of the one\hyp{}particle density matrix susceptibilities (DMS)
interacting with the spatial distribution of electric field.
The analytical approximate expressions are derived
at the Hartree\hyp{}Fock theory, which serves as a basis for construction of generalized
model that can be used for an arbitrary form of wavefunction as long as the one\hyp{}particle
density matrices are available. It is shown that it is possible to accurately predict the changes
in the one\hyp{}electron ground\hyp{}state density of water molecule in spatially uniform electric field, as well as
in spatially non\hyp{}uniform electric field distribution
generated by point charges. When both linear and quadratic terms with respect to the electric field 
are accounted for,
the electric field\hyp{}induced polarization energies,
dipole moments and quadrupole moments are quantitatively described by the present theory 
in electric fields from weak to very strong in magnitude (0.003 -- 0.07 a.u.).
It is believed that the proposed model could open
new routes in Quantum Chemistry for fast and efficient calculations of molecular properties
in condensed phases.
\end{abstract}

@@

Most of the efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wider scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,\cite{Holas.March.PhysRevA.1991}
though obtaining the exact density has been an undersolved issue for most of chemical systems.
Given that fact, it would be incredibly insightful 
to establish a first\hyp{}principles connection between the one\hyp{}particle
density and the perturbation in the environment through introduction of
\emph{density matrix susceptibility} (DMS) tensors. 
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%
\begin{equation*} \label{e:initial-model}
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential which in principle, is applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
because the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.

The simplest type of perturbation, yet the most relevant for chemistry, 
is the electrostatic perturbation.
It is worth mentioning that, at the moment, there is no simple \emph{ab initio}
method to compute electron deformation densities without considering the entire molecular 
aggregate by using full quantum chemistry calculations.\cite{Horn.Head-Gordon.JCP.2015,
Mandado.Hermida-Ramon.JCTC.2011}
In this work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion 
or dispersion,\cite{Mandado.Hermida-Ramon.JCTC.2011}
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.

@@

In the limit of a weak and spatially uniform electric field,
the polarization\hyp{}induced density matrix change is given by the follwoing Taylor expansion:
%
\begin{equation}\label{e:dD-Taylor}
 \delta D_{\alpha\beta} = \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}}  \cdot {\bf F} 
   + \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} : {\bf F} \otimes {\bf F}
   + \ldots 
\end{equation}
%
In the above expression, the density matrix susceptibilities are simply equal to the derivatives of the density matrix
with respect to electric field (up to a multiplicative constant).
In a general case, i.e., when electric field is non\hyp{}uniform, the development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
becomes extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element will accumulate when performing two\hyp{}
and four\hyp{}index contractions,
and in turn, affect the evaluated properties with non\hyp{}negligible errors.
Single\hyp{}centered expansions, like the one from Eq.~\eqref{e:dD-Taylor}, although valid for 
an arbitrary level of theory,
are not sufficient enough to accurately describe the local polarization
of electron density distribution when electric fields are highly non\hyp{}uniform in space.
Due to all of these reasons, we adopt the following approach: 
we first develop a qualitative first\hyp{}principles distributed\hyp{}site model at the Hartree\hyp{}Fock
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that
%
\begin{equation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{M}} \cdot \delta {\bf C}
   \right] }  
-
2 \Tr{ 
    \left[ 
         \widetilde{\mathbb{K}} \cdot \delta {\bf C}
   \right] } \;.
\end{equation}
%
Here we introduced the auxiliary tensors
%
\begin{subequations}
 \begin{align}
   \widetilde{\mathbb{M}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}}     \;,           \\
   \widetilde{\mathbb{K}}  &\equiv {\bf C}^{(0)\dagger} \cdot {\mathbb{M}} \cdot {\bf D}^{(0)} \;, \\
   \widetilde{\mathbb{L}}  &\equiv \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \;
 \end{align}
\end{subequations}
%
to simplify the notation. Hence, we have
%
\begin{equation} \label{e:dmu-l-vector}
  \frac{1}{4} 
 \delta {\BM{\upmu}} %({\bf r}_0) 
   =
   \Tr{ 
    \left[ 
         \widetilde{\mathbb{L}} \cdot \delta {\bf C}
    \right] }
   %
   \equiv \sum_{i}^{\rm occ} \sum_\alpha {\bf l}_{i\alpha} \delta C_{\alpha i} \;,
\end{equation}
%
where the Cartesian vector
%
\begin{equation}\label{e:l-vector}
 {\bf l}_{i\alpha} \equiv \left[ \widetilde{\mathbb{M}} - \widetilde{\mathbb{K}} \right]_{i\alpha} 
      = \left[  {\bf C}^\dagger \cdot \mathbb{M} \cdot \left( {\bf 1} - {\bf D}^{(0)} \right) \right]_{i\alpha}
\end{equation}
%
and $i$ runs over occupied unperturbed MO's.
Note that the intermediate quantity $\widetilde{\mathbb{L}}$ contains the projector onto
the unoccupied MO space at the absence of the external electric field. It means that the
perturbed solution is constructed from the unperturbed virtual orbitals as well.
Although Eq.~\eqref{e:dmu-l-vector}
is exact up to first order in ${\BM\Delta}$, it cannot be inverted at this moment to obtain $\delta {\bf C}$
due to the presence of trace operation, unless the electric field is uniform.
Let the induced dipole moment be partitioned into separate contributions
from the occupied MO's, i.e.,
%
\begin{equation} \label{e:dmu-part}
 \delta {\BM{\upmu}} = \sum_i^{\rm occ} \delta {\BM{\upmu}}_i \;,
\end{equation}
%
where
%
\begin{equation} \label{e:dmu-dpol}
 \delta {\BM{\upmu}}_i = {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) \;.
\end{equation}
%
In Eq.~\eqref{e:dmu-dpol}, ${\BM{\alpha}}_i$ is the distributed dipole\hyp{}dipole polarizability tensor
associated with the $i$th MO whereas ${\bf F}({\bf r}_i)$ is the electric field evaluated at the $i$th MO centroid of charge,
${\bf r}_i = \tbraket{\phi_i}{\hat{\bf r}}{\phi_i}$.
By inserting the above result into Eq.~\eqref{e:dmu-l-vector},
recasting the right hand side in a matrix form
and assuming real quantities,
one can separate the contributions coming from occupied MO's as
%
\begin{equation} \label{e:dmu-l-vector-mo-transform-X}
 \frac{1}{4} {\BM{\alpha}}_i \cdot {\bf F}({\bf r}_i) 
   =
   \delta {\bf c}_i^T \cdot {\bf L}_i \;.
\end{equation}
%
Note that now ${\bf L}_i$ is a matrix of dimension $n \times 3$ and $\delta{\bf c}_i$ is the $i$th column of the
change in the LCAO-MO matrix due to the polarization process.
The solution for $\delta {\bf C}$ is
%
\begin{equation} \label{e:mu-ind.matrix.linear.solution}
  \delta {\bf c}_i^T = \frac{1}{4}
            {\bf F}({\bf r}_i)  \cdot {\BM{\alpha}}_i \cdot 
                    \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \;,
\end{equation}
%
where $\left[ {\bf L}_i  \right]^{-1}_{\rm Left}$ is a left inverse
of ${\bf L}_i$ matrix,
%
\begin{equation} 
      \left[ {\bf L}_i  \right]^{-1}_{\rm Left}   \equiv
       \left[ {\bf L}_i^T \cdot {\bf L}_i \right]^{-1} \cdot {\bf L}_i^T   
\end{equation}
%
with the dimension of $3\times n$. Note that the right inverse of ${\bf L}_i$
does not exist because ${\bf L}_i \cdot {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

Eq.~\eqref{e:final-model.HF}, which is the final result of the previous section, 
can be used as a guideline to construct a general model of density matrix polarization.
Quadratic (and higher\hyp{}order) terms with respect to the electric field
can be added to the formulation, which leads to
%
\begin{multline}\label{e:final-model.General}
 \delta D_{\alpha\beta} = \sum_{i }^M \Big\{
                                      {\bf B}^{(i ; 1)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \\
%                       + \sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)
                        +             {\bf B}^{(i ; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) 
                        + \ldots \Big\} \;.
\end{multline}
%
In the above equation, the density matrix susceptibilities
${\bf B}^{(i; R)}_{\alpha\beta}$, distributed over $M$ sites,
denote the $R$-th order response with respect to the electric field.
Note that
${\bf B}^{(i; 1)}_{\alpha\beta}$ is a vector of length $3$ whereas ${\bf B}^{(i; 2)}_{\alpha\beta}$
is a $3\times 3$ matrix. Also, the distributed expansion centres
(labeled as $i$) are generalized to any point in the molecule.\footnote{More general
contribution due to square of electric field could 
be $\sum_{ij}^M {\bf B}^{(ij; 2)}_{\alpha\beta} : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_j)$. 
However, for simplicity, we neglect the contributions from different sites $i$ and $j$
in the present discussion and leave it for future development.
}

@@

The exact mathematical form of the generalized density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown. 
The susceptibility from Eq.~\eqref{e:final-model.HF} is valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarizability approximation), 
but not higher multipole moments and polarization energy, are correctly described.
Our strategy is to treat the susceptibilities from Eq.~\eqref{e:final-model.General}
as adjustable parameters, rather than try to derive them from the first principles theory.

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule
was selected as a test system which was subjected to 
electrostatic perturbations due to random uniform electric fields and random collections of point charges
generating non\hyp{}uniform electric fields. 
Components of spatially uniform electric fields were determined 
from the uniform distribution in the interval of $\vert 0.0404 \vert$ a.u.
generating average field of 0.035 a.u. in an ensemble of 100 samples.
Cartesian coordinates of point charges were drawn from the uniform distribution 
of a set of fourty three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
with an origin that is located in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Electric charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly).
In this way, independent sets of 100 random samples of uniform electric fields as well as 
40 point charges were selected and exact solutions of HF\hyp{}SCF equations were 
performed for each validation procedure.

@@

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;, 
  \end{align}
\end{subequations}
%
where the subscript `CM' denotes the center of mass of water molecule.
When the distributed polarizabilities were used instead of the polarization\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments (given in Eq.~\eqref{e:dmu-dpol}) according to
%
\begin{equation}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
Methodology for the calculation of the electric field due to point charges
can be found elsewhere.\cite{Blasiak.Cho.JCP.2014}
